---
title: "Simulating Multivariate Normal by Univariate Normal"
date: 2017-12-25T17:03:40-06:00
draft: true
author: "Yanyu Liang"
tags: ['simulation', 'multivariate normal', 'contour']
categories: ["rscript"]
---

```{r setup}
library(knitr)
opts_chunk$set(tidy.opts=list(width.cutoff=60),tidy=TRUE)
```

# Univariate to Multivariate

\begin{align}
  \vec{y} = \sum_i \vec{v}_i \mathcal{N}_i \nonumber
\end{align}

, where $\mathcal{N}_1, ..., \mathcal{N}_p$ are independent standard univariate normal. Then

\begin{align}
  \vec{y} \sim \mathcal{N}(\vec{0}, VV^T) \label{eq:mvn}
\end{align}

, where $V = [\vec{v}_1, ..., \vec{v}_p]$.

# Example

In the following example, $v_i \in \mathbb{R}^2$ and $p = 2$. The number of simulated samples is $B = 10000$.

## Ploting contour in R

```{r}
# draw contour of a multivariate normal distribution
library(mvtnorm)
drawContour <- function(xgrid, ygrid, mu, sigma) {
  z <- c()
  x <- c()
  y <- c()
  for(i in 1 : length(xgrid)) {
    for(j in 1 : length(ygrid)) {
      z <- c(z, dmvnorm(c(xgrid[i], ygrid[j]), mean = mu, sigma = sigma))
      x <- c(x, xgrid[i])
      y <- c(y, ygrid[j])
    }
  }
  return(data.frame(x = x, y = y, z = z))
}
```

## Simulating and ploting

```{r}
library(MASS)
library(ggplot2)

# simulation
v1 <- c(1, 2)
v2 <- c(-1, -1)
V <- matrix(c(v1, v2), ncol = 2, byrow = F)
B <- 1000 # number of simulated samples
p <- 2 # number of vectors
n1 <- rnorm(B)
n2 <- rnorm(B)
n <- matrix(c(n1, n2), ncol = 2, byrow = F)
y <- t(V %*% t(n))
df.samples <- data.frame(x = y[, 1], y = y[, 2])

# ploting - contour
xgrid <- seq(from = min(y[, 1]) - 0.1, max(y[, 1]) + 0.1, length.out = 100)
ygrid <- seq(from = min(y[, 2]) - 0.1, max(y[, 2]) + 0.1, length.out = 100)
mu <- rep(0, p)
sigma <- V %*% t(V)
df <- drawContour(xgrid, ygrid, mu, sigma)
ggplot() + geom_contour(data = df, aes(x = x, y = y, z = z, alpha = ..level..)) + geom_point(data = df.samples, aes(x = x, y = y), alpha = 0.1)

# ploting - heatmap
y.mvn <- mvrnorm(B, rep(0, p), V %*% t(V))
y.combine <- rbind(y, y.mvn)
label <- c(rep('uni -> multi',nrow(y)), rep('multivariate normal', nrow(y.mvn)))
df <- data.frame(x = y.combine[, 1], y = y.combine[, 2], label = label)
ggplot(df) + geom_bin2d(aes(x = x, y = y)) + facet_grid(. ~ label)
```

# Proof

This is one of the definition of multivariate normal distribution according to [wikipedia](https://en.wikipedia.org/wiki/Multivariate_normal_distribution#Definition). Here, suppose $\vec{y}$ is multivariate normal and let's check the mean and variance is the same as the one in $\eqref{eq:mvn}$.

$$
\newcommand\E{\text{E}}
\newcommand\var{\text{Var}}
\newcommand\cov{\text{Cov}}
\newcommand\nocr{\nonumber\\}
\def\ci{\perp\!\!\!\!\perp}
$$

\begin{align}
  \E(\vec{y}) &= \E(\sum_i v_i Z_i) \text{, where $Z_i \sim \mathcal{N}(0, 1)$} \nocr
  &= \sum_i \E(v_i)\E(Z_i) \text{, since $v_i \ci Z_i$} \nocr
  &= 0 \nocr
  \cov(y_k, y_j) &= \cov(\sum_i v_{ik} Z_i, \sum_{i'} v_{i'j} Z_{i'}) \nocr
  &= \sum_{i}\sum_{i'}\cov(v_{ik}Z_i, v_{i'j}Z_{i'}) \nocr
  &= \sum_i \cov(v_{ik}Z_i, v_{ij}Z_{i}) \text{, since $v_i \ci v_{i'}$ and $Z_i \ci Z_{i'}$} \nocr
  &= \sum_i v_{ik}v_{ij} \nocr
  \therefore \cov(\vec{y}) &= V V^T \nonumber
\end{align}
